\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-09-23 doc/paper-jss/sdmTMB-paper.Rnw
Software section work

The assumption that correlation decays equally in all directions can be relaxed to allow for geometric anisotropy
\[
\Phi\left( s_j, s_k \right) =
  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \bm{H} \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \bm{H} \right).
\]
Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{lindgren2011, thorson2015ices, thorson2018comp}.

@@

%GMRF
%is an SPDE whose solution exhibits the desired covariance structure and associated precision matrix.
%The approach outlined in \citet{lindgren2011} streamlines the computation of Q and he demonstrates that
%Can then take advantage of methods for rapid precision matrix computation \citep{rue2005}

%The SPDE approach offers an efficient means to estimate Q without the typical computational overhead. It revolves around solving an equation known as the SPDE. The solutions to this equation are stochastic processes, and their covariance structure is tailored to meet the specifications outlined by the SPDE. To employ the SPDE approach effectively for f, the goal is to identify an SPDE whose solutions exhibit the desired covariance structure and implied precision matrix. In the work of Lindgren et al. (2011), a practical solution to the SPDE is presented by representing f as a combination of basis functions multiplied by coefficients. This approach significantly streamlines the computation of Q. Moreover, Lindgren et al. (2011) demonstrate that these coefficients form a Gaussian Markov random field, benefiting from existing methods for swift precision matrix computation (Rue and Held 2005). These computations pave the way for rapid model fitting using integrated nested Laplace approximations (INLA; Rue et al. 2009).

%The SPDE approach links

%A user should know that...
%A GMRF has the advantage that it can be fit modelling the precision matrix, which is computationally much more efficient than working with the covariance matrix.

%Because the parameters and input matrices are revelant to understanding sdmTMB internals, we note that the precision

%a way to approximate Q
%SPDE = an equation to be solved

%We introduce the SPDE approach to working with the precision matrix in Section \ref{sec:software}, but for now, note that we can instead represent our spatial random field as
%\[
%\boldsymbol{\omega} \sim
%  \mathrm{MVNormal}\left(\boldsymbol{0}, \boldsymbol{Q}^{-1}_\omega\right),
%\]
%where $\boldsymbol{Q}_\omega$ is a sparse precision matrix.
%%
% The SPDE approach does just that, ... have been developed lets us stay with precision and so as precision:
%
% The approach involves finding an SPDE w

% Details are beyond the scope of this paper and are not necessary to use the
% software discussed here, but the idea is that the solution to a specific SPDE is
% a GRF with a Matérn covariance function and this `trick' enables one to
% efficiently fit approximations to GRFs to large spatial datasets
% \citep{lindgren2011}.

% "piece-wise linear basis functions defined by a triangulation of the domain of interest"

% Miller et al: The SPDE approach provides a method to approximate Q without the common substantial computational burden. The SPDE is an equation to be solved. Solutions to this equation are stochastic processes whose covariance structure is chosen to satisfy the relationship the SPDE specifies. The SPDE approach involves finding an SPDE whose solutions have the covariance structure, and implied precision matrix, that is desired for f . Lindgren et al. (2011) show how to find an approximate solution to the SPDE by representing f as a sum of basis functions multiplied by coefficients; this provides a computationally efficient way to compute Q: Lindgren et al. (2011) show that the coefficients of these basis functions form a Gaussian Markov random field, for which methods for fast computation of precision matrices already exist (Rue and Held 2005). These computations make it possible to fit models quickly using integrated nested Laplace approximations (INLA; Rue et al. 2009).

@@

In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors. All these components are additive in link space. Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a more complex model could be written as
\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\alpha_g +
O_{\bm{s},t} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right).
\end{aligned}
\]
where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}). We demonstrate these model components in the examples below.

@@

So far we have described models with one linear predictor and one family (common terminology in \proglang{R} packages for the combination of an observation likelihood and link). Frequently, data are better represented with two-part ``delta'' or ``hurdle'' models, which include linear predictors and observation distributions for two processes: zero vs.\ non-zero values and positive values, respectively \citep{aitchison1955}. Here, we describe two types of delta models, dropping the space and time subscripts for simplicity.
% and use $\bm{X_1} \bm{\beta_1}$ and $\bm{X_2} \bm{\beta_2}$ to refer to generic model matrices and parameter vectors for the first and second model components.

@@

\section{break}

\pkg{sdmTMB} fits GLMMs to spatial or spatiotemporal geostatistical data.
Geostatistical data refers to data observed at specific spatial coordinates
reflecting some underlying spatial process \citep{rossi1992, diggle2007}. These
data can be collected across discrete points in time. The process component of
an \pkg{sdmTMB} model can be formed by any combination of main (``fixed'')
effects \(\bm{\beta}\), an offset \(O_{\bm{s},t}\), IID (independent and
identically distributed) random intercepts \(\alpha_g\), spatial intercept
random fields \(\omega_{\bm{s}}\), spatiotemporal intercept random fields
\(\epsilon_{\bm{s},t}\), time-varying effects \(\bm{\gamma_t}\), and spatially
varying effects \(\zeta_{\bm{s}}\) (Figure~\ref{fig:diagram}). These components
are discussed in more detail via our examples below.

This process component is combined with an observation error family (e.g.,
Gaussian, gamma, binomial, Tweedie) and link function \(g\) (e.g., identity,
log, logit) as in any generalized linear model (GLM) \citep{mccullagh1989}. Some
families can be combined into a two-part ``delta'' or ``hurdle'' model
\citep{aitchison1955} to model the zero vs.~non-zero observations separately
from the positive observations. See the model description vignette for an
expanded description of the statistical model
(\texttt{browseVignettes("sdmTMB")} or see the
\href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).

The GLMMs underpinning \pkg{sdmTMB} models are spatially explicit---they
estimate interpretable parameters of a spatial covariance function: parameters
defining the magnitude of spatial variation and the rate of correlation decay
with distance. In contrast, semi- and non-parametric approaches do not estimate
spatial covariance functions (e.g., \pkg{randomForest} \citep{liaw2002},
\pkg{MaxEnt} \citep{phillips2006}, and most smooths in \pkg{mgcv}
\citep{wood2017a}). The random fields in \pkg{sdmTMB} are structured as MVN
constrained by a Matérn covariance function \citep{matern1960}. The Matérn can
accommodate a range of shapes and can be both isotropic (correlation decays the
same in all directions) or anisotropic (correlation is directionally dependent)
\citep{haskard2007}. The Matérn standard deviations are estimated separately for
the various fields and the range---the distance at which spatial correlation
decays to \(\sim 0.13\) \citep{lindgren2015}---can be shared among random fields
or estimated separately (\texttt{share\_range} argument).

TODO INTEGRATE THIS SOMEWHERE: Areal data (data aggregated to polygon or grid
level) may be analyzed using other spatial models, including conditional (CAR)
autoregressive models \citep[e.g.,][]{verhoef2018}; \pkg{sdmTMB} can also fit
models with areal data if each polygon has an associated centroid. A benefit of
the geostatistical approach over CAR or similar models is that the parameters
describing spatial covariance can be more easily interpreted \citep{wall2004}
(e.g., spatial variance and range---the distance at which points are effectively
independent).

By default, if spatiotemporal fields are included, they are assumed IID;
however, additional options allow them to be modelled as a random walk or
first-order autoregressive, AR(1), process (Figure~\ref{fig:diagram}). Turning
off both spatial and spatiotemporal effects allows for comparison with standard
non-spatial GLMs or GLMMs. We include additional flexibility in specifying the
linear main effect matrix: covariates can be modelled as penalized smooth
functions (generalized additive models, GAMs) using the same \texttt{s()} syntax
as in \pkg{mgcv} \citep{wood2017a} (thereby allowing automatic selection of
smoother `wiggliness'), or can be modelled with threshold shapes: hockey-stick
models, \texttt{breakpt()}, \citep{barrowman2000} or logistic functions,
\texttt{logistic()}. \pkg{sdmTMB} can use many \texttt{mgcv::s()} and
\texttt{mgcv::t2()} smoothers including bivarate smoothers \texttt{s(x, y)},
smoothers varying by continuous or categorical variables \texttt{s(x1, by =
x2)}, cyclical smoothers \texttt{s(x, bs = "cc")}, and smoothers with
specified basis dimensions \texttt{s(x, k = 4)} \citep{wood2017a}.

\pkg{sdmTMB} models can include penalized likelihoods by assigning priors to
model parameters (\texttt{?sdmTMBpriors}). These priors may be useful in cases
where estimation is difficult because of identifiability issues or relatively
flat likelihood surfaces, or to impart prior information or achieve
regularization. Following other recent SPDE implementations in \pkg{TMB}
\citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors
\citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the
spatial range and variance parameters.

@@

The design goal of \sdmTMB\ was to make...

\begin{figure}[p]
\centering
\includegraphics[width=4in]{flowchart}
\caption[]{
TODO
}\label{fig:flowchart}
\end{figure}

\sdmTMB\ relies on several well-established \R\ packages (FIG?).
\sdmTMB\ uses \pkg{fmesher} and \INLA\ to generate input matrices for SPDE calculations \citep[$\bm{C}$, $\bm{G}$, and $\bm{K}$ in][]{lindgren2011}.
The SPDE calculations themselves are performed by functions within \pkg{TMB} (\texttt{R\_inla} namespace).
Data structures required to implement penalized smoothers are constructed with \pkg{mgcv}'s \texttt{smooth2random()} function similarly to \pkg{gamm4} and \pkg{brms}.
The statistical model itself is written as a \proglang{C++} template for \TMB.
\TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to estimate the joint posterior distribution for Bayesian inference.

Is modular in nature...

INLA/fmesher - TMB - R combination - describe here.
+ tmbstan

Describe modular design philosopy

The design goal of...

flexible, modular, familiar, make SPDE accessible and fast

combine functionality commonly needed together

The \pkg{sdmTMB} model is fit by maximum marginal likelihood. Internally,
a \pkg{TMB} \citep{kristensen2016} model template is used to calculate the
marginal log likelihood and its gradient, and the negative log likelihood is
minimized via the non-linear optimization routine \texttt{stats::nlminb()} in
\proglang{R} \citep{gay1990, r2021}. Random effects are estimated at values that
maximize the log likelihood conditional on the estimated fixed effects and are
integrated over via the Laplace approximation \citep{kristensen2016}. After
rapid model exploration with maximum likelihood, one can optionally pass an
\pkg{sdmTMB} model to the \proglang{R} package \pkg{rstan} \citep{carpenter2017,
rstan} via \pkg{tmbstan} \citep{monnahan2018} to estimate the joint posterior
distribution for Bayesian inference.

The \pkg{sdmTMB} package is constructed from S3 classes and designed to be both
modular and familiar to users of widely used \proglang{R} packages (e.g.,
\pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}). Data may be simulated with or without
a fitted model (\texttt{simulate.sdmTMB()}, \texttt{sdmTMB\_simulate()}
respectively). Estimation is done via maximum marginal likelihood with the
\texttt{sdmTMB()} function, and the resulting fitted model can be used to make
predictions (\texttt{predict.sdmTMB()}) or calculate residuals
(\texttt{residuals.sdmTMB()}), derive additional quantities of interest, and
simulate Bayesian posterior distributions. Common quantities of interest may
include summarizing shifts in the center of gravity (\texttt{get\_cog()})
\citep[e.g.,][]{thorson2016cog}, or tracking total densities through time
(\texttt{get\_index()}). The \texttt{get\_index()} function generates a weighted
average of densities for the prediction grid, integrating across space; for
prediction grids with equally sized cells, each cell will contribute equally to
the total average, however cells may be of unequal size.
% TODO: describe this in more detail and how it could apply beyond ecology;
% maybe a derived variables subsection? Expanded this sentence a bit -- COGs are
% pretty common in other fields, but I don't know about indices -- removed
% `ecology` from the description

% \begin{landscape}
\begin{figure}[p]
\centering
\includegraphics[width=\textwidth]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples,
notation, and example code. An \pkg{sdmTMB} model can be built from any
combination of the process components (first six rows) plus an observation
component (last row). The examples are from an SDM context, but the model can be
fit to any spatially referenced point data. Notation: We refer to design
matrices as $\bm{X}$. The indexes $\bm{s}$, $t$, and $g$ index spatial
coordinates, time, and group, respectively. The $\sigma$ and $\bm{\Sigma}$
symbols represent standard deviations and covariance matrices, respectively. All
other symbols refer to the  described model components (e.g., $\bm{\beta}$ and
$\bm{\omega}$ refer to a vector of main effects and spatial random field
deviations, respectively). See the \pkg{sdmTMB}
\href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette}
for a full description of the model. Note that \code{s()} denotes a smoother as
in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint
`hockey-stick' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes
a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in
an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
% \end{landscape}

@@

\section{Model validation and selection}\label{model-validation-and-selection}

Validation and selection of state-space models is challenging, particularly when
using the Laplace approximation \citep{thygesen2017a}. We provide several
approaches to assist this process: (1) The Akaike Information Criterion
\citep[AIC,][]{akaike1974} can be calculated with \texttt{AIC()}, although AIC
has well-documented biases with mixed-effects models \citep{liang2008}. (2)
Alternatively, k-fold cross validation with \texttt{sdmTMB\_cv()} can be used
with user-specified or randomly chosen folds for model selection or to evaluate
goodness of fit according to user-calculated criteria (e.g., mean squared error,
area under the curve). (3) An \pkg{sdmTMB} model can be passed to the
\pkg{tmbstan} package \citep{monnahan2018} to sample from the joint posterior
with Stan \citep{carpenter2017}, evaluate the accuracy of the Laplace
approximation, or perform posterior predictive checks (see
\texttt{?extract\_mcmc}). (4) The \texttt{residuals()} method by default returns
randomized quantile \citep{dunn1996} or probability integral transform (PIT)
\citep{smith1985} residuals. For state-space models, these residuals have known
statistical issues with the Laplace approximation \citep{thygesen2017a} but are
quick to calculate. A version that uses Markov chain Monte Carlo (MCMC) sampling
of the random effects to avoid this issue \citep{rufener2021} is recommended but
slower (\texttt{?residuals.sdmTMB}). Simulation-based residuals \citep{dharma}
are also possible. (5) The \texttt{simulate.sdmTMB()} method can simulate from
fitted models and the \texttt{sdmTMB\_simulate()} function facilitates
simulating data without starting from a fitted model. Models can be fit to these
simulated data to ensure identifiability, evaluate bias and precision in
parameter estimation, or evaluate the consequences of model misspecification.

\section{Installation}\label{installation}

@@

Additional utilities are maintained in the \pkg{sdmTMBextra} package at
\url{https://github.com/pbs-assess/sdmTMBextra}.

@@

\sloppy We then create a mesh object that contains triangulation and projection
matrices needed to apply the SPDE approach using \texttt{make\_mesh()}.
\texttt{make\_mesh()} is a wrapper function for functions in the \pkg{fmesher}
package. The argument \texttt{cutoff} defines the minimum allowed distance
between mesh vertices in the units of \texttt{X} and \texttt{Y} (km). We could create a basic mesh just specifying this:

@@

We can use the \texttt{tidy()} function \citep[formatted to match the
\pkg{broom} package;][]{broom} to obtain a data frame with parameter estimates
(standard methods such as \texttt{fixef()}, \texttt{confint()}, and
\texttt{vcov()} are also available). The standard errors on our fixed effects
have increased with the spatial random field:

>>>>

The assumption that correlation decays equally in all directions can be relaxed to allow for geometric anisotropy
\[
\Phi\left( s_j, s_k \right) =
  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \bm{H} \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \bm{H} \right).
\]
Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{haskard2007, lindgren2011, thorson2015ices, thorson2018comp}.

@@

@@

In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors (Figure~\ref{fig:diagram}). All these components are additive in link space. Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a more complex model could be written as
\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\alpha_g +
O_{\bm{s},t} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right).
\end{aligned}
\]
where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}). We demonstrate these model components in the examples below.
% TODO: add MVN(0, sigma)s

\begin{figure}[p]
\centering
\includegraphics[width=\textwidth]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples,
notation, and example code. An \pkg{sdmTMB} model can be built from any
combination of the process components (first six rows) plus an observation
component (last row). The examples are from an SDM context, but the model can be
fit to any spatially referenced point data. Notation: We refer to design
matrices as $\bm{X}$. The indexes $\bm{s}$, $t$, and $g$ index spatial
coordinates, time, and group, respectively. The $\sigma$ and $\bm{\Sigma}$
symbols represent standard deviations and covariance matrices, respectively. All
other symbols refer to the  described model components (e.g., $\bm{\beta}$ and
$\bm{\omega}$ refer to a vector of main effects and spatial random field
deviations, respectively). See the \pkg{sdmTMB}
\href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette}
for a full description of the model. Note that \code{s()} denotes a smoother as
in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint
`hockey-stick' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes
a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in
an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}

@@

So far we have described models with one linear predictor and one family (common terminology in \R\ packages for the combination of an observation likelihood and link). Frequently, data are better represented with two-part ``delta'' or ``hurdle'' models, which include linear predictors and observation distributions for two processes: zero vs.\ non-zero values and positive values, respectively \citep{aitchison1955}. Here, we describe two types of delta models, dropping the space and time subscripts for simplicity.
% and use $\bm{X_1} \bm{\beta_1}$ and $\bm{X_2} \bm{\beta_2}$ to refer to generic model matrices and parameter vectors for the first and second model components.

@@

% \pkg{sdmTMB} fits GLMMs to spatial or spatiotemporal geostatistical data.
% Geostatistical data refers to data observed at specific spatial coordinates
% reflecting some underlying spatial process \citep{rossi1992, diggle2007}. These
% data can be collected across discrete points in time. The process component of
% an \pkg{sdmTMB} model can be formed by any combination of main (``fixed'')
% effects \(\bm{\beta}\), an offset \(O_{\bm{s},t}\), IID (independent and
% identically distributed) random intercepts \(\alpha_g\), spatial intercept
% random fields \(\omega_{\bm{s}}\), spatiotemporal intercept random fields
% \(\epsilon_{\bm{s},t}\), time-varying effects \(\bm{\gamma_t}\), and spatially
% varying effects \(\zeta_{\bm{s}}\) (Figure~\ref{fig:diagram}). These components
% are discussed in more detail via our examples below.

% This process component is combined with an observation error family (e.g.,
% Gaussian, gamma, binomial, Tweedie) and link function \(g\) (e.g., identity,
% log, logit) as in any generalized linear model (GLM) \citep{mccullagh1989}. Some
% families can be combined into a two-part ``delta'' or ``hurdle'' model
% \citep{aitchison1955} to model the zero vs.~non-zero observations separately
% from the positive observations. See the model description vignette for an
% expanded description of the statistical model
% (\texttt{browseVignettes("sdmTMB")} or see the
% \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).

% By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Figure~\ref{fig:diagram}). 
% Turning off both spatial and spatiotemporal effects allows for comparison with standard non-spatial GLMs or GLMMs.

@@

The design goal of \sdmTMB\ was to develop a flexible, modular, and familiar inteface to fast maximum likelihood inference (or full Bayesian inference) with the SPDE approach to spatial and spatiotemporal GLMMs with random fields.
The package gathers functionality not found combined in other packages that is particularly useful to species distribution modelling, but is applicable far beyond ecology.

\sdmTMB\ relies on several well-established \R\ packages (Figure~\ref{fig:flowchart}).
\sdmTMB\ uses \pkg{fmesher} \citep{fmesher} and \INLA\ \citep{lindgren2015} to generate the triangulation mesh and input matrices for the SPDE calculations.
The SPDE precision matrix calculations themselves are performed by functions within \pkg{TMB} \citep{kristensen2016} (\texttt{R\_inla} namespace).
Data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
The statistical model itself is written as a \proglang{C++} template for \TMB.
\TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.

\begin{figure}[htbp]
\centering
\includegraphics[width=3.8in]{flowchart}
\caption[]{
TODO
}\label{fig:flowchart}
\end{figure}

The \sdmTMB\ package is designed to be both modular and familiar to users of widely used \R\ packages (e.g., \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}). 
The user starts by constructing a triangulation mesh for the SPDE approach with \texttt{make\_mesh()}. 
\texttt{make\_mesh()} is a wrapper for \pkg{fmesher} and \INLA's triangulation mesh functions and users can also construct any mesh with \pkg{fmesher} or \INLA\ and pass it to \texttt{make\_mesh()}.
Fitting is accomplished with \texttt{sdmTMB()}, which has arguments similar to \pkg{glmmTMB}'s \texttt{glmmTMB()} but with additional arguments for how any spatial and spatiotemporal random fields are structured, what column defines time, any time-varying formulas, and any spatially varying formulas.
Families are specified with standard R family functions (e.g., \texttt{binomial()}) plus several \sdmTMB-specific families for family's not available in the \R\ stats library (e.g., \texttt{nbinom2()}, \texttt{delta\_lognormal()}).
The formula syntax follows a standard \R\ format and is expanded on in the next paragraph.
A fitted model summary can be viewed with \texttt{print.sdmTMB()} or \texttt{summary.sdmTMB()} and a set of basic ``sanity'' checks can be run with \texttt{sanity()}.
\texttt{tidy()} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
Others standard S3 methods are also available such as \texttt{fixef()}, \texttt{confint()}, and \texttt{vcov()}.
Prediction on fitted or new data is accomplished with \texttt{predict.sdmTMB()}.
\texttt{residuals.sdmTMB()} calculates various types of residuals.
New observations can be simulated from a fitted model with \texttt{simulate.sdmTMB()} or new data can be simulated without a fitted model (``de novo'') with \texttt{sdmTMB\_simulate()}.
\texttt{sdmTMB\_cv()} facilitates cross validation.
Derived quantities can be calculated on the fitted model.
% TODO: more COG refs
Common quantities of interest include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}). 
% The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.

% Data may be simulated with or without a fitted model (\texttt{simulate.sdmTMB()}, \texttt{sdmTMB\_simulate()} respectively). 
% Estimation is done via maximum marginal likelihood with the \texttt{sdmTMB()} function, and the resulting fitted model can be used to make predictions (\texttt{predict.sdmTMB()}) or calculate residuals (\texttt{residuals.sdmTMB()}), derive additional quantities of interest, and simulate Bayesian posterior distributions. 
% Common quantities of interest may include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}). 
% The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.

The \sdmTMB\ formula syntax (\texttt{formula}) follows a standard format that is similar to \pkg{glmmTMB}, \pkg{lme4}, and \pkg{mgcv}.
In addition to standard main effects, the user can include random intercepts (e.g., \texttt{+ (1 | group)}), threshold shaped hockey-stick models \texttt{breakpt()} \citep{barrowman2000}, logistic functions \texttt{logistic()}, and smoothers.
Penalized smoothers (i.e., for generalized additive models, GAMs) use the same \texttt{s()} and \texttt{t2()} syntax as in \pkg{mgcv} \citep{wood2017a}.
Supported functionality includes bivarate smoothers \texttt{s(x, y)},
smoothers varying by continuous or categorical variables \texttt{s(x1, by = x2)},
cyclical smoothers \texttt{s(x, bs = "cc")},
and smoothers with specified basis dimensions \texttt{s(x, k = 4)} \citep{wood2017a}.
In addition to the main formula, \texttt{sdmTMB()} accepts \emph{one-sided} formulas for coefficients that should vary through time (\texttt{time\_varying}) according to a random walk or AR(1) process (\texttt{time\_varying\_type}) or vary through space as random fields (\texttt{spatial\_varying}).

The \texttt{predict.sdmTMB()} method is flexible and includes the option to specify a new data frame (\texttt{newdata}),
whether to return predictions on the link or response scale (type),
whether to return standard errors (\texttt{se\_fit},
can be slow if conditioned on random fields),
whether to condition on the random fields (\texttt{re\_form}),
whether to condition on the random intercepts (\texttt{re\_form\_iid}),
which delta model linear predictor to use (\texttt{model}),
whether to take draws from the joint parameter precision matrix (\texttt{nsim}),
and whether to use MCMC samples from a \pkg{tmbstan} model fit (\texttt{mcmc\_samples}).

\pkg{sdmTMB} models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \texttt{sdmTMB()} \texttt{prior} argument (\texttt{?sdmTMBpriors}). 
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization. 
Following other recent SPDE implementations in \pkg{TMB} \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the spatial range and variance parameters.
These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan}, a \texttt{bayesian} logical argument is available to enable Jacobian adjustments for change of variables (priors applied to parameters that are internally transformed) (REF).

% Describe modular design philosopy

% The design goal of...

% flexible, modular, familiar, make SPDE accessible and fast

% combine functionality commonly needed together

% The \pkg{sdmTMB} model is fit by maximum marginal likelihood. Internally,
% a \pkg{TMB} \citep{kristensen2016} model template is used to calculate the
% marginal log likelihood and its gradient, and the negative log likelihood is
% minimized via the non-linear optimization routine \texttt{stats::nlminb()} in
% \proglang{R} \citep{gay1990, r2021}. Random effects are estimated at values that
% maximize the log likelihood conditional on the estimated fixed effects and are
% integrated over via the Laplace approximation \citep{kristensen2016}. After
% rapid model exploration with maximum likelihood, one can optionally pass an
% \pkg{sdmTMB} model to the \proglang{R} package \pkg{rstan} \citep{carpenter2017,
% rstan} via \pkg{tmbstan} \citep{monnahan2018} to estimate the joint posterior
% distribution for Bayesian inference.

% TODO: describe this in more detail and how it could apply beyond ecology;
% maybe a derived variables subsection? Expanded this sentence a bit -- COGs are
% pretty common in other fields, but I don't know about indices -- removed
% `ecology` from the description

@@

% \section{Model validation and selection}\label{model-validation-and-selection}

% Validation and selection of state-space models is challenging, particularly when
% using the Laplace approximation \citep{thygesen2017a}. We provide several
% approaches to assist this process: (1) The Akaike Information Criterion
% \citep[AIC,][]{akaike1974} can be calculated with \texttt{AIC()}, although AIC
% has well-documented biases with mixed-effects models \citep{liang2008}. (2)
% Alternatively, k-fold cross validation with \texttt{sdmTMB\_cv()} can be used
% with user-specified or randomly chosen folds for model selection or to evaluate
% goodness of fit according to user-calculated criteria (e.g., mean squared error,
% area under the curve). (3) An \pkg{sdmTMB} model can be passed to the
% \pkg{tmbstan} package \citep{monnahan2018} to sample from the joint posterior
% with Stan \citep{carpenter2017}, evaluate the accuracy of the Laplace
% approximation, or perform posterior predictive checks (see
% \texttt{?extract\_mcmc}). (4) The \texttt{residuals()} method by default returns
% randomized quantile \citep{dunn1996} or probability integral transform (PIT)
% \citep{smith1985} residuals. For state-space models, these residuals have known
% statistical issues with the Laplace approximation \citep{thygesen2017a} but are
% quick to calculate. A version that uses Markov chain Monte Carlo (MCMC) sampling
% of the random effects to avoid this issue \citep{rufener2021} is recommended but
% slower (\texttt{?residuals.sdmTMB}). Simulation-based residuals \citep{dharma}
% are also possible. (5) The \texttt{simulate.sdmTMB()} method can simulate from
% fitted models and the \texttt{sdmTMB\_simulate()} function facilitates
% simulating data without starting from a fitted model. Models can be fit to these
% simulated data to ensure identifiability, evaluate bias and precision in
% parameter estimation, or evaluate the consequences of model misspecification.

\subsection{Installation}\label{installation}

@@

Additional utilities are maintained in the \pkg{sdmTMBextra} package at\\
\url{https://github.com/pbs-assess/sdmTMBextra}

@@

\sloppy We then create a mesh object that contains triangulation and projection
matrices needed to apply the SPDE approach using \texttt{make\_mesh()}.
% \texttt{make\_mesh()} is a wrapper function for functions in the \pkg{fmesher} package. 
The argument \texttt{cutoff} defines the minimum allowed distance
between mesh vertices in the units of \texttt{X} and \texttt{Y} (km). 
We could create a basic mesh just specifying this:

@@

We can use the \texttt{tidy()} function to obtain a data frame with parameter estimates (standard methods such as \texttt{fixef()}, \texttt{confint()}, and \texttt{vcov()} are also available). 
The standard errors on our fixed effects have increased with the spatial random field:
